Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-08-20 paper.Rnw
Half of Matt's comments responded to

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Sweave/Knitr init
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw(base_size = 12))

@@

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, url, geometry, ctable}
\usepackage{titlesec, siunitx}
\usepackage[round,authoryear]{natbib}
\usepackage[section]{placeins}  % keep floats in their place.
\usepackage{authblk}            % Listing Author affiliations
\usepackage[hypcap]{caption}

@@

% shortcuts 
\newcommand{\mb}{MAGeT-Brain }
\newcommand{\ants}{ANTS }
\newcommand{\animal}{ANIMAL }
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T }
\newcommand{\FEPdataset}{SZ-FEP }

@@

Automated segmentation techniques that are reliable, objective, and reproducible
are a necessary alternative to manual segmentation.  In the case of classical
model-based segmentation methods \citep{Haller1997,Csernansky1998}, an MRI atlas
that was previously manually labelled by an expert rater is matched to target
images using nonlinear registration methods.  The resulting nonlinear
transformation is applied to the manual labels (i.e. {\em label propagation}) to
move them into the target image space.  While this methodology has been used
successfully in several contexts
\citep{Chakravarty2008,Chakravarty2009,Collins1995,Haller1997}, it is limited
in accuracy by the introduction of errors due inaccuracies in the nonlinear
transformation itself, partial volume effects in label resampling, and
irreconcilable differences between the neuroanatomy represented within the atlas
and target images.

@@

More recently, many groups have been using multiple atlases to improve overall
segmentation accuracy (i.e. multi-atlas segmentation) over model-based approaches
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2010,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used to
merge these labels into a definitive segmentation for the target.  In addition,
weighted voting procedures that use {\em atlas selection} techniques are often
used to exclude atlases from label fusion that are dissimilar to a target image
in order to reduce error from unrepresentative anatomy \citep{Aljabar2009}.
This involves the selection of a subset of atlases using a similarity metric
such as cross-correlation \citep{Aljabar2009} or normalized mutual information.
Such selection has the added benefit of significantly reducing the number of
nonlinear registrations.  For example \citet{Collins2010}
demonstrated that only 14 atlases, selected based on highest similarity between
medial temporal lobe neuroanatomy as evaluated by normalized mutual information
\citep{Studholme1999} from a library of 80 atlases, were required to achieve
accurate segmentations of the hippocampus.  Several methods have been explored
for label fusion including the STAPLE algorithm \citep{Warfield2004} that
computes a probabilistic segmentation using an expectation maximization
framework from an set of competing segmentations, or others where a subset of
segmentations can be estimated using metrics such as the sum of squared
differences in the regions of interest to be segmented \citep{Coupe2011}.

However, many of these methods require significant investment of time and
resources for the creation of the atlas library ranging between 30
\citep{Heckemann2006} and 80 \citep{Collins2010} manually segmented atlases.
This strategy has the main drawback of being inflexible as it does not easily
accommodate varying the definition of the hippocampal anatomy (such as the
commonly used heuristic of subdividing the hippocampus into head, body, and tail
\citep{Poppenk2011,Pruessner2000}).  Furthermore, none of these methods have
demonstrated sufficient flexibility to accommodate atlases that are somehow
exceptional such as those derived from serial histological data
\citep{Chakravarty2006,Yelnik2007} or high-resolution MRI data that enables
robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}.  Due
to the recent availability of the latter, there has been increased interest in
the use of probabilistic methods for the identification of the hippocampal
subfields on standard T1-weighted images.  Our group recently demonstrated that
through use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data is possible \citep{MallarChakravarty2012}.
The novelty of this manuscript is the extension our multi-atlas methodology to
the hippocampus using more than a single input atlas, while simultaneously
limiting the number of possible inputs used during segmentation, and
demonstrating that accurate identification of the hippocampal subfields is
indeed possible using this methodology.

There are few methods that have attempted to perform multi-atlas segmentation
with a limited number of input atlases. The LEAP algorithm is an elegant
modification to the basic multi-atlas strategy \citep{Wolz2010} in which the
atlas library is grown, beginning with a set of manually labelled atlases, and
successively incorporating unlabelled target images after themselves being
labelling using multi-atlas techniques. The sequence in which target images are
labelled is chosen so that the similarity between the atlas images and the
target images is minimised at each step, effectively allowing for deformations
between very dissimilar images to be broken up into sequences of smaller
deformations.  Although \citet{Wolz2010} begin with an atlas library of 30 MR
images, this method could theoretically work using a much smaller atlas library.
In their validation, LEAP was used to segment the whole hippocampus in the
ADNI-1 baseline dataset, achieving a mean Dice score of 0.85 with manual
segmentations.

To the best of our knowledge there are two other segmentation methods that
attempt to define the hippocampal subfields using standard T1-weighted data.
The first is included with the FreeSurfer package \citep{VanLeemput2009}.  This
work is limited as it omits the tail of the hippocampus and the segmentation
protocol has yet to be fully validated.  Nonetheless, it  demonstrates that the
applicability of hippocampal subfield segmentation using data from 10 subjects.
In the second method, \citet{Yushkevich2009} acquired and labelled hippocampal
subfields on high-resolution MRI data from post-mortem medial temporal lobe
samples. Using nonlinear registration guided by manually derived hippocampus
masks and specific landmarks, they demonstrate accurate parcellation of
hippocampal subfields in unlabelled MRI volumes.

@@

The simplest form of multi-atlas segmentation, (which we will call {\em basic 
multi-atlas segmentation}), involves three steps.  First, each labelled 
input image (i.e. atlas or template) is registered to an unlabelled target image.
Second, the labels from each image are propagated to the target image space. 
Third, the labels are combined into a single labelling by way of a label fusion method 
\citep{Heckemann2006, Heckemann2011}. This method is described in detail in other
publications \citep{Collins2010,Heckemann2011,Aljabar2009}.

\mb bootstraps the creation of a large template library given a limited input atlas 
library, and then uses the template library in basic multi-atlas segmentation. 
Images for the template library are selected from a set of input target images, 
either arbitrarily or so as to reflect the neuroanatomy or demographics of the 
target set as a whole (for instance, by sampling equally from cases and controls). 
The template library images are then labelled by each of the atlases. Basic 
multi-atlas segmentation is then conducted using the template library
to segment the entire set of target images (including the targets whose images
are used in the construction of the template library).  Since each template
library image has multiple labels (one from each atlas), the final number of 
labels to be fused for each target may be quite large (i.e. \# of atlas $\times$ 
\# of templates).

@@

The following section describes the experiments we conducted to assess the
segmentation quality of the \mb algorithm. The first two experiments assess the
validity of \mb using a cross-validation design. Experiment 1 investigates the
accuracy of \mb whole hippocampus segmentation over a wide range of parameter
settings. The results of this experiment enable us to choose the parameter
settings offering the best performance for use in subsequent experiments.
Experiment 2 tests hippocampal subfield segmentation quality. The last two
experiments assess the validity of the \mb algorithm when applied to different
diseases: first episode schizophrenia (Experiment 3), and Alzheimer's disease
(Experiment 4). Additionally, in experiment 4, we compare \mb segmentations with
those of well-known automated and manual methods and assess segmentation bias.

@@

Monte Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using a pool
of images and manual hippocampal segmentations from ADNI1 dataset.  This form of
cross-validation allows us to rigorously validate a large number of parameter
settings of \mb (atlas and template library sizes, registration algorithm, and 
label fusion method) and select the best parameters to use in subsequent 
experiments.

\paragraph{ADNI1 dataset}

69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset} standardized dataset. 23 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Manual segmentations of the left and
right whole hippocampi are available \citep{Hsu2002}. These labels have been
generated using the SNT tool from Medtronic Surgical Navigation Technologies,
Louisville, CO (see Supplementary Materials for detailed discussion of the
manual segmentation process used).

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset download was the "\adnidataset"
standardized dataset available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly preprocessed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008}.  Representative 1.5T imaging parameters were TR
= 2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240
x 240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel resolution of $1.25 \times 1.25 \times 1.2mm^3$.

@@

In this experiment, our dataset consists of 69 1.5T images and corresponding
manual segmentations. In each validation round, the dataset is partitioned into 
a training set consisting of images and their manual labels to be used as an atlas 
library, and a validation set consisting of the remaining images segmented by both 
\mb and multi-atlas. The resulting segmentations are compared to the manual 
segmentations for the images.

@@

Before registration, all images underwent preprocessing with the N3 algorithm
\citep{Sled1998} to minimize intensity nonuniformity. In this experiment we use
one of two non-linear image registration methods.

@@

The \animal algorithm carries out image registration is in two phases. In the
first, a 12-parameter linear transformation (3 translations, rotations, scales,
shears) is estimated between images using an algorithm that maximizes the
correlation between blurred MR intensities and gradient magnitude over the whole
brain \citep{Collins}. In the second phase, nonlinear registration is completed
using the \animal algorithm \citep{Collins1995}: an iterative procedure that
estimates a 3D deformation field between two MR images. At first, large
deformations are estimated using blurred version of the input data. These larger
deformations are then input to subsequent steps where the fit is refined by
estimating smaller deformations on data blurred with a Gaussian kernel with a
smaller full width at half maximum (FWHM). The final transformation is a set of
local translations defined on a bed of equally spaced nodes that were estimated
through the optimization of the correlation coefficient.  For the purposes of
this work we used the regularization parameters optimized in
\citet{Robbins2004}, displayed in Table \ref{tab:ANIMAL-params}.

@@

\paragraph{Label fusion methods}
Label fusion is a term given to the process of combining the information from
several candidate labellings for an intensity image into a single labelling.  In
this experiment we explore three fusion methods. 
\begin{description}
  \item[Voxel-wise Majority Vote]
  Labels are propagated from all template library images to a target.  Each
  output voxel is given the most frequent label at that voxel location amongst
  all candidate labellings.  Ties are broken arbitrarily.

@@

  \item[Normalised Mutual Information Weighted Majority Vote]
  This method is similar to cross-correlation weighted voting except that image
  similarity is calculated by the normalised mutual information score over the
  region of interest\citep{Studholme2001}.  The {\tt itk\_similarity} utility
  from the EZMinc toolkit\footnote{\url{https://github.com/vfonov/EZminc}} is
  used to calculate the normalised mutual information measure between to images.
\end{description}
\paragraph{Evaluation method}

The Dice similarity coefficient (DSC) assesses the agreement between two
segmentations. It is one of the most widely used measures of segmentation
performance, and we use it as the basis of comparison in this experiment.
%Additionally, we report the Jaccard index, another commonly used similarity
%measure:

@@

The manual segmentations (SNT) provided as part of the ADNI dataset are used as
the gold standard with which to compare \citep{Hsu2002}. The segmentation
accuracy reported is averaged over the ten validation rounds for each parameter
setting.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%        Experiment 2: Hippocampal Subfield Cross-Validation          %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 2: Hippocampal Subfield Cross-Validation}
In this experiment, the accuracy of the \mb algorithm on hippocampal subregion
segmentation is tested using a leave-one-out cross-validation (LOOCV) design. 
The optimal parameter settings for \mb found in Experiment 1 are used in this 
experiment.

@@

In addition to the high-resolution scans distributed as part of the Winterburn
atlases, we also obtained additional 3T T1 BRAVO images (0.9mm-isotropic voxels)
of four of the five Winterburn atlas subjects.

\paragraph{Experiment details} 
Leave-one-out cross-validation (LOOCV) is an approach in which the method to be
validated is given all but one item in a dataset as training data, and the
output is compared with the left-out item. This is done, in turn, for each item
in the dataset.

In this experiment the Winterburn atlases are used as the cross-validation
dataset.  The five $300\mu$-isotropic voxel images and labels are used as
atlases, and LOOCV is conducted using the five Winterburn atlas subject images
subsampled (using trilinear interpolation) to 0.9mm-isotropic voxel resolution
(referred to as the {\em Subsampled} dataset) as input subjects. Subsampling the
subject images allows us to assess \mb in a typical segmentation scenario
(high-resolution atlases and lower-resolution subjects).  The template library
is composed of the subject images, plus an additional set of 3T T1 images
(0.9mm-isotropic voxels) of healthy subjects acquired separately (Table
\ref{tab:WAval-healthy-demographics}). The optimal size of template library,
registration method, and label fusion method found in Experiment 1 are used.
Each resampled Winterburn atlas subject image is segmented by \mb with that
subject's image excluded from the atlas library.

@@

We reproduce this experiment in a separate LOOCV run in which the resampled 
Winterburn atlas subject images are substituted for separately acquired T1 
BRAVO images of four of the subjects (referred to as the {\em BRAVO} dataset).

@@

To validate that \mb algorithm works effectively in the context of other
neuropsychiatric disorders, we use the Winterburn atlases with \mb to predict
the hippocampal segmentation of dataset of Schizophrenia patient MR images. The
resulting segmentations are assessed for quality by comparison with expert
manual segmentations.

@@

Each subject hippocampus is traced following a validated segmentation protocol 
\citep{Pruessner2000}.

@@

\paragraph{Evaluation method}
The manual and Winterburn hippocampal segmentation protocols differ slightly in
the neuroanatomical features that are delineated. This difference poses a
problem for evaluation by measuring overlap.  That is, since different protocols
will necessarily produce segmentations that do not perfectly overlap, the degree
of overlap cannot be solely used to compare segmentation methods using different
protocols. In place of an overlap metric, we can assess the degree of
correlation in average hippocampal volume of the subjects produced by each
method. Specifically, a Pearson correlation is used.

@@

To validate the \mb algorithm as respect to other established automated
methods, \mb is applied to large dataset from the ADNI project, and the
resulting segmentations are compared to those produced by FreeSurfer, FSL,
MAPER, and by expert semi-automated segmentation (SNT).

@@

\paragraph{Evaluation method}
As in Experiment 3, the manual (SNT) and Winterburn hippocampal segmentation
protocols differ in the neuroanatomical features delineated, and so we must
assess \mb by the degree of correlation of average hippocampal volume across all
subjects produced by \mb and by manual segmentation.  Specifically, Pearson
correlation is used.  For comparison, we also compute the correlation in
hippocampal volume between the existing, established automated segmentation
methods: FSL, FreeSurfer, and MAPER.  Additionally, we evaluate the
volume-related fixed and proportional biases in all segmentation methods using
Bland-Altman plots \citep{Bland1986}.

@@

In this experiment \mb is applied to a dataset of images of first episode
schizophrenia patients, using the Winterburn atlases and a template library of
21 subject images selected at random.  Expert manual whole hippocampal
segmentations are used as gold standards.

@@

We found a close relationship in total bilateral hippocampal volume between
all methods and manually segmented volumes (Figure \ref{fig:ADNI-volumes-plot}). 
Volumes are correlated with Pearson $r > 0.78$ for all methods across disease
categories.  Within disease categories (Figure \ref{fig:ADNI-volumes-boxplot}),
\mb is consistently well correlated to manual volumes (Pearson $r > 0.85$), but
appears to slightly over-estimate the volume of the AD hippocampus.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has 
%  - update montages with TAL coordinates
%  - add MAGeT brain diagram, and remove pseudocode
%  - check that \gs is used instead of Manual 
% Sweave/Knitr init
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw(base_size = 12))

@@

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, url, geometry, ctable}
\usepackage{titlesec, siunitx, graphicx}
\usepackage[round,authoryear]{natbib}
\usepackage[section]{placeins}  % keep floats in their place.
\usepackage{authblk}            % Listing Author affiliations
\usepackage[hypcap]{caption}

@@

% shortcuts 
\newcommand{\mb}{MAGeT-Brain }
\newcommand{\ants}{ANTS }
\newcommand{\animal}{ANIMAL }
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T }
\newcommand{\FEPdataset}{SZ-FEP }
\newcommand{\gs}{Gold Standard }

@@

Automated segmentation techniques that are reliable, objective, and reproducible
are a necessary alternative to manual segmentation. In the case of classical
model-based segmentation methods \citep{Haller1997,Csernansky1998}, an MRI atlas
that was previously manually labelled by an expert rater is matched to target
images using nonlinear registration methods. The resulting nonlinear
transformation is applied to the manual labels (i.e. {\em label propagation}) to
move them into the target image space. While this methodology has been used
successfully in several contexts
\citep{Chakravarty2008,Chakravarty2009,Collins1995,Haller1997}, it is limited in
accuracy due to inaccuracies in the estimated nonlinear transformation itself,
partial volume effects in label resampling, and irreconcilable differences
between the neuroanatomy represented within the atlas and target images.

@@

More recently, many groups have been using multiple atlases to improve overall
segmentation accuracy (i.e. multi-atlas segmentation) over model-based approaches
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2010,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used to
merge these labels into the definitive segmentation for the target. In addition,
weighted voting procedures that use {\em atlas selection} techniques are often
used to exclude atlases from label fusion that are dissimilar to a target image
in order to reduce error from unrepresentative anatomy \citep{Aljabar2009}.
This involves the selection of a subset of atlases using a similarity metric
such as cross-correlation \citep{Aljabar2009} or normalized mutual information.
Such selection has the added benefit of significantly reducing the number of
nonlinear registrations. For example \citet{Collins2010} demonstrated that only
14 atlases, selected based on highest similarity between medial temporal lobe
neuroanatomy as evaluated by normalized mutual information \citep{Studholme1999}
from a library of 80 atlases, were required to achieve accurate segmentations of
the hippocampus. Additionally, several methods have been explored for label
fusion, including the STAPLE algorithm \citep{Warfield2004} that computes a
probabilistic segmentation using an expectation maximization framework from an
set of competing segmentations, or others where a subset of segmentations can be
estimated using metrics such as the sum of squared differences in the regions of
interest to be segmented \citep{Coupe2011}.

However, many of these methods require significant investment of time and
resources for the creation of the atlas library ranging between 30
\citep{Heckemann2006} and 80 \citep{Collins2010} manually segmented atlases.
This strategy has the main drawback of being inflexible as it does not easily
accommodate varying the definition of the hippocampal anatomy (such as the
commonly used heuristic of subdividing the hippocampus into head, body, and tail
\citep{Poppenk2011,Pruessner2000}). Furthermore, none of these methods have
demonstrated sufficient flexibility to accommodate atlases that are somehow
exceptional such as those derived from serial histological data
\citep{Chakravarty2006,Yelnik2007} or high-resolution MRI data that enables
robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}. Due
to the recent availability of the latter, there has been increased interest in
the use of probabilistic methods for the identification of the hippocampal
subfields on standard T1-weighted images. Our group recently demonstrated that
through use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data is possible \citep{MallarChakravarty2012}.
The novelty of this manuscript is the extension of our multi-atlas methodology
to the hippocampus using more than a single input atlas, while simultaneously
limiting the number of possible inputs used during segmentation, and
demonstrating that accurate identification of the hippocampal subfields is
indeed possible using this methodology.

There are few methods that have attempted to perform multi-atlas segmentation
with a limited number of input atlases. The LEAP algorithm is an elegant
modification to the basic multi-atlas strategy \citep{Wolz2010} in which the
atlas library is grown, beginning with a set of manually labelled atlases, and
successively incorporating unlabelled target images after themselves being
labelled using multi-atlas techniques. The sequence in which target images are
labelled is chosen so that the similarity between the atlas images and the
target images is minimised at each step, effectively allowing for deformations
between very dissimilar images to be broken up into sequences of smaller
deformations. Although \citet{Wolz2010} begin with an atlas library of 30 MR
images, this method could theoretically work using a much smaller atlas library.
In their validation, LEAP was used to segment the whole hippocampus in the
ADNI-1 baseline dataset, achieving a mean Dice score of 0.85 with manual
segmentations.

To the best of our knowledge there are two other segmentation methods that
attempt to define the hippocampus and its subfields using standard T1-weighted
data. The first is included with the FreeSurfer package \citep{VanLeemput2009}.
This work is limited as it omits the tail of the hippocampus and the
segmentation protocol has yet to be fully validated. Nonetheless, it
demonstrates that the applicability of hippocampal subfield segmentation using
data from 10 subjects. In the second method, \citet{Yushkevich2009} acquired
and labelled hippocampal subfields on high-resolution MRI data from post-mortem
medial temporal lobe samples. Using nonlinear registration guided by manually
derived hippocampus masks and specific landmarks, they demonstrate accurate
parcellation of hippocampal subfields in unlabelled MRI volumes.

@@

The simplest form of multi-atlas segmentation, (which we will call {\em basic 
multi-atlas segmentation}), involves three steps. First, each labelled 
input image (i.e. atlas or template) is registered to an unlabelled target image.
Second, the labels from each image are propagated to the target image space. 
Third, the labels are combined into a single label by label fusion
\citep{Heckemann2006, Heckemann2011}. This method is described in detail in
other publications \citep{Collins2010,Heckemann2011,Aljabar2009}.

\mb bootstraps the creation of a large template library given a limited input atlas 
library, and then uses the template library in basic multi-atlas segmentation. 
Images for the template library are selected from a set of input target images, 
either arbitrarily or so as to reflect the neuroanatomy or demographics of the 
target set as a whole (for instance, by sampling equally from cases and controls). 
The template library images are automatically labelled by each of the atlases
via label propagation. Basic multi-atlas segmentation is then conducted using
the template library to segment the entire set of target images (including the
targets whose images are used in the construction of the template library).
Since each template library image has multiple labels (one from each atlas), the
final number of labels to be fused for each target may be quite large (i.e. \#
of atlas $\times$ \# of templates).

@@

The following section describes for experiments conducted to assess the
segmentation quality of the \mb algorithm. The first two experiments assess the
validity of \mb using a cross-validation design. Experiment 1 investigates the
accuracy of \mb whole hippocampus segmentation over a wide range of parameter
settings. The results of this experiment enable us to choose the parameter
settings offering the best performance for use in subsequent experiments.
Experiment 2 tests hippocampal subfield segmentation quality. The last two
experiments assess the validity of the \mb algorithm when applied to different
diseases: first episode schizophrenia (Experiment 3), and Alzheimer's disease
(Experiment 4). Additionally, in experiment 4, we compare \mb segmentations with
those of well-known automated and manual methods and assess segmentation bias.

@@

Monte Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using a pool
of images and SNT \gs segmentations from the ADNI1 dataset. This form of
cross-validation allows us to rigorously validate a large number of parameter
settings of \mb (atlas and template library sizes, registration algorithm, and
label fusion method) and select the best parameters to use in subsequent
experiments.

\paragraph{\adnidataset dataset} 
69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset} standardized dataset. 23 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Semi-automated segmentations of the
left and right whole hippocampi are made available with a subset of ADNI images
\citep{Hsu2002}. These labels have been generated using the SNT tool from
Medtronic Surgical Navigation Technologies, Louisville, CO (see Supplementary
Materials for detailed discussion of the segmentation process used). Throughout
this paper, we will use these segmentations as the \gs for comparison.

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset used was the "\adnidataset" standardized dataset
available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly preprocessed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008}.  Representative 1.5T imaging parameters were TR =
2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240 x
240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel resolution of $1.25 \times 1.25 \times 1.2mm^3$.

@@

In this experiment, our dataset consists of 69 1.5T images and corresponding \gs
segmentations. In each validation round, the dataset is partitioned into a
training set consisting of images and their \gs segmentations to be used as an
atlas library, and a validation set consisting of the remaining images segmented
by both \mb and multi-atlas. The computed segmentations are compared to the
\gs segmentations (see below).

@@

Before registration, all images underwent preprocessing with the N3 algorithm
\citep{Sled1998} to minimize intensity nonuniformity. In this experiment we
compared two non-linear image registration methods:

@@

The \animal algorithm carries out image registration in two phases. In the
first, a 12-parameter linear transformation (3 translations, rotations, scales,
shears) is estimated between images using an algorithm that maximizes the
correlation between blurred MR intensities and gradient magnitude over the whole
brain \citep{Collins}. In the second phase, nonlinear registration is completed
using the \animal algorithm \citep{Collins1995}: an iterative procedure that
estimates a 3D deformation field between two MR images. At first, large
deformations are estimated using blurred version of the input data. These larger
deformations are then input to subsequent steps where the fit is refined by
estimating smaller deformations on data blurred with a Gaussian kernel with a
smaller full width at half maximum (FWHM). The final transformation is a set of
local translations defined on a bed of equally spaced nodes that were estimated
through the optimization of the correlation coefficient.  For the purposes of
this work we used the regularization parameters optimized in
\citet{Robbins2004}, displayed in Table \ref{tab:ANIMAL-params}.

@@

\paragraph{Label fusion methods}
Label fusion is a term given to the process of combining the information from
several candidate labellings for an intensity image into a single labelling.  In
this experiment we explore three fusion methods. 
\begin{description}
  \item[Voxel-wise Majority Vote]
  Labels are propagated from all template library images to a target.  Each
  output voxel is given the most frequent label at that voxel location amongst
  all candidate labellings.

@@

  \item[Normalised Mutual Information Weighted Majority Vote]
  This method is similar to cross-correlation weighted voting except that image
  similarity is calculated by the normalised mutual information score over the
  region of interest\citep{Studholme2001}.  The {\tt itk\_similarity} utility
  from the EZMinc toolkit\footnote{\url{https://github.com/vfonov/EZminc}} is
  used to calculate the normalised mutual information measure between to images.
\end{description}

\paragraph{Evaluation method}

The Dice similarity coefficient (DSC), also known as Dice's Kappa, assesses the
agreement between two segmentations. It is one of the most widely used measures
of segmentation accuracy, and we use it as the basis of comparison in this
experiment.
%Additionally, we report the Jaccard index, another commonly used similarity
%measure:

@@

The automatically produces labels are compared to the \gs SNT labels provided by
ADNI. The segmentation accuracy reported for a parameter setting is the average
over ten validation rounds.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%        Experiment 2: Hippocampal Subfield Cross-Validation          %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 2: Hippocampal Subfield Cross-Validation}
In this experiment, the accuracy of the \mb algorithm on hippocampal subregion
segmentation is tested using a leave-one-out cross-validation (LOOCV) design on
two image sets. The optimal parameter settings for \mb found in Experiment 1 are
used in this experiment.

@@

In addition to the high-resolution scans distributed as part of the Winterburn
atlases, we also obtained separate 3T T1 BRAVO acquisitions (0.9mm-isotropic
voxels) of four of the five Winterburn atlas subjects.

\paragraph{Experiment details} 
Leave-one-out cross-validation (LOOCV) is an approach in which the method to be
validated is given all but one item in a dataset as training data (in our case,
atlas library images and labels), and the output accuracy is measured on the
left-out item. This is done, in turn, for each item in the dataset.

\begin{figure}
  \begin{centering}
    \includegraphics[width=\textwidth]{figure/WinterburnSubfieldLOOCVSchematic}
  \end{centering}
  \caption{A schematic illustration of the experimental set up of the
  leave-one-out cross-validation (LOOCV) subfield experiment.}
  \label{fig:LOOCV-schematic}
\end{figure}

In part 2a of this experiment, we have modified the LOOCV experimental
design in order to compare \mb segmentation accuracy with the accuracy of image
resampling error. The five $300\mu$-isotropic voxel Winterburn atlas images and
labels are used as the atlas library, and LOOCV is conducted using the five Winterburn
atlas images subsampled (using trilinear interpolation) to
0.9mm-isotropic voxel resolution (referred to as the {\em Subsampled} dataset)
as input subjects. Subsampling the subject images allows us to assess \mb in a
typical segmentation scenario (high-resolution atlases and lower-resolution
subjects).  The template library is composed of all five subject images, plus an
additional set of 3T T1 images (0.9mm-isotropic voxels) of healthy subjects
acquired separately (Table \ref{tab:WAval-healthy-demographics}). The optimal
size of template library, registration method, and label fusion method found in
Experiment 1 are used.  As with traditional LOOCV, each subject image is
segmented by \mb with that subject's image excluded from the atlas library.  In
part 2b of this experiment, we carry out a similar validation in which the
subject images are separately acquired T1 BRAVO images of four of the five
Winterburn atlas subjects (referred to as the {\em BRAVO} dataset).  Figure
\ref{fig:LOOCV-schematic} illustrates the experimental set up of both parts of
this experiment.

@@

@@

To validate that \mb works effectively in the context of other neuropsychiatric
disorders, in this experiment we use the Winterburn atlases to predict the
hippocampal segmentations of patients with Schizophrenia. The resulting
segmentations are assessed for quality by comparison with expert manual
segmentations.

@@

Manual segmentation of each subject hippocampus is produced following a
validated segmentation protocol \citep{Pruessner2000}.

@@

\paragraph{Evaluation method}
The manual and Winterburn hippocampal segmentation protocols differ slightly in
the neuroanatomical features that are delineated \citep{Winterburn2013}. This
difference poses a problem for evaluation by measuring overlap.  That is, since
different protocols will necessarily produce segmentations that do not perfectly
overlap, the degree of overlap cannot be solely used to compare segmentation
methods using different protocols. In place of an overlap metric, we can assess
the degree of correlation in average hippocampal volume of the subjects produced
by each method. Specifically, a Pearson correlation is used.

@@

To validate the \mb algorithm with respect to other established automated
methods, \mb is applied to large dataset from the ADNI project and the
resulting segmentations are compared to those produced by FreeSurfer, FSL,
MAPER and by expert semi-automated manual segmentations (SNT).

@@

\paragraph{Evaluation method}
As in Experiment 3, the SNT \gs and Winterburn hippocampal segmentation
protocols differ in the neuroanatomical features delineated, and so we must
assess \mb by the degree of correlation of average hippocampal volume across all
subjects. Specifically, Pearson correlation is used. We also compute the
correlation in hippocampal volume between the existing, established automated
segmentation methods: FSL, FreeSurfer, and MAPER.  Additionally, we evaluate the
volume-related fixed and proportional biases in all segmentation methods using
Bland-Altman plots \citep{Bland1986}.

@@

In this experiment \mb is applied to a dataset of images of first episode
schizophrenia patients, using the Winterburn atlases and a template library of
21 subject images selected at random.  Expert manual whole hippocampal
segmentations are used as a gold standard.

@@

We found a close relationship in total bilateral hippocampal volume between all
methods and SNT \gs volumes (Figure \ref{fig:ADNI-volumes-plot}).  Volumes are
correlated with Pearson $r > 0.78$ for all methods across disease categories.
Within disease categories (Figure \ref{fig:ADNI-volumes-boxplot}), \mb is
consistently well correlated to manual volumes (Pearson $r > 0.85$), but appears
to slightly over-estimate the volume of the AD hippocampus.
